---
title: "`AMR` with `tidymodels`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{`AMR` with `tidymodels`}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE, results = 'markup'}
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.5,
  fig.height = 5
)
```

Antimicrobial resistance (AMR) is a global health crisis, and understanding resistance patterns is crucial for managing effective treatments. The `AMR` R package provides robust tools for analysing AMR data, including convenient antibiotic selector functions like `aminoglycosides()` and `betalactams()`. In this post, we will explore how to use the `tidymodels` framework to predict resistance patterns in the `example_isolates` dataset. 

By leveraging the power of `tidymodels` and the `AMR` package, we’ll build a reproducible machine learning workflow to predict resistance to two important antibiotic classes: aminoglycosides and beta-lactams.

---

### **Objective**

Our goal is to build a predictive model using the `tidymodels` framework to determine resistance patterns based on microbial data. We will:

1. Preprocess data using the selector functions `aminoglycosides()` and `betalactams()`.
2. Define a logistic regression model for prediction.
3. Use a structured `tidymodels` workflow to preprocess, train, and evaluate the model.

---

### **Data Preparation**

We begin by loading the required libraries and preparing the `example_isolates` dataset from the `AMR` package.

```{r}
# Load required libraries
library(tidymodels)   # For machine learning workflows, and data manipulation (dplyr, tidyr, ...)
library(AMR)          # For AMR data analysis

# Load the example_isolates dataset
data("example_isolates")  # Preloaded dataset with AMR results

# Select relevant columns for prediction
data <- example_isolates %>%
  # select AB results dynamically
  select(mo, aminoglycosides(), betalactams()) %>%
  # replace NAs with NI (not-interpretable)
   mutate(across(where(is.sir),
                 ~replace_na(.x, "NI")),
          # make factors of SIR columns
          across(where(is.sir),
                 as.integer),
          # get Gramstain of microorganisms
          mo = as.factor(mo_gramstain(mo))) %>%
  # drop NAs - the ones without a Gramstain (fungi, etc.)
  drop_na() # %>%
  # Cefepime is not reliable
  #select(-FEP)
```

**Explanation:**
- `aminoglycosides()` and `betalactams()` dynamically select columns for antibiotics in these classes.
- `drop_na()` ensures the model receives complete cases for training.

---

### **Defining the Workflow**

We now define the `tidymodels` workflow, which consists of three steps: preprocessing, model specification, and fitting.

#### 1. Preprocessing with a Recipe

We create a recipe to preprocess the data for modelling. This includes:
- Encoding resistance results (`S`, `I`, `R`) as binary (resistant or not resistant).
- Converting microbial organism names (`mo`) into numerical features using one-hot encoding.

```{r}
# Define the recipe for data preprocessing
resistance_recipe <- recipe(mo ~ ., data = data) %>%
  step_corr(c(aminoglycosides(), betalactams()), threshold = 0.9)
resistance_recipe
```

**Explanation:**
- `step_mutate()` transforms resistance results (`R`) into binary variables (TRUE/FALSE).
- `step_dummy()` converts categorical organism (`mo`) names into one-hot encoded numerical features, making them compatible with the model.

#### 2. Specifying the Model

We define a logistic regression model since resistance prediction is a binary classification task.

```{r}
# Specify a logistic regression model
logistic_model <- logistic_reg() %>%
  set_engine("glm") # Use the Generalized Linear Model engine
logistic_model
```

**Explanation:**
- `logistic_reg()` sets up a logistic regression model.
- `set_engine("glm")` specifies the use of R's built-in GLM engine.

#### 3. Building the Workflow

We bundle the recipe and model together into a `workflow`, which organizes the entire modeling process.

```{r}
# Combine the recipe and model into a workflow
resistance_workflow <- workflow() %>%
  add_recipe(resistance_recipe) %>% # Add the preprocessing recipe
  add_model(logistic_model) # Add the logistic regression model
resistance_workflow
```

---

### **Training and Evaluating the Model**

To train the model, we split the data into training and testing sets. Then, we fit the workflow on the training set and evaluate its performance.

```{r}
# Split data into training and testing sets
set.seed(123) # For reproducibility
data_split <- initial_split(data, prop = 0.8) # 80% training, 20% testing
training_data <- training(data_split) # Training set
testing_data <- testing(data_split)   # Testing set

# Fit the workflow to the training data
fitted_workflow <- resistance_workflow %>%
  fit(training_data) # Train the model

fitted_workflow
```

**Explanation:**
- `initial_split()` splits the data into training and testing sets.
- `fit()` trains the workflow on the training set.

Next, we evaluate the model on the testing data.

```{r}
# Make predictions on the testing set
predictions <- fitted_workflow %>%
  predict(testing_data)                # Generate predictions
probabilities <- fitted_workflow %>%
  predict(testing_data, type = "prob") # Generate probabilities

predictions <- predictions %>%
  bind_cols(probabilities) %>%
  bind_cols(testing_data) # Combine with true labels

predictions

# Evaluate model performance
metrics <- predictions %>%
  metrics(truth = mo, estimate = .pred_class) # Calculate performance metrics

metrics
```

**Explanation:**
- `predict()` generates predictions on the testing set.
- `metrics()` computes evaluation metrics like accuracy and AUC.

It appears we can predict the Gram based on AMR results with a `r round(metrics$.estimate[1], 3)` accuracy. The ROC curve looks like:

```{r}
predictions %>%
  roc_curve(mo, `.pred_Gram-negative`) %>%
  autoplot()
```

---

### **Conclusion**

In this post, we demonstrated how to build a machine learning pipeline with the `tidymodels` framework and the `AMR` package. By combining selector functions like `aminoglycosides()` and `betalactams()` with `tidymodels`, we efficiently prepared data, trained a model, and evaluated its performance.

This workflow is extensible to other antibiotic classes and resistance patterns, empowering users to analyse AMR data systematically and reproducibly.

---
