---
title: "AMR with tidymodels"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{AMR with tidymodels}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE, results = 'markup'}
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.5,
  fig.height = 5
)
```

> This page was entirely written by our [AMR for R Assistant](https://chatgpt.com/g/g-M4UNLwFi5-amr-for-r-assistant), a ChatGPT manually-trained model able to answer any question about the AMR package.

Antimicrobial resistance (AMR) is a global health crisis, and understanding resistance patterns is crucial for managing effective treatments. The `AMR` R package provides robust tools for analysing AMR data, including convenient antimicrobial selector functions like `aminoglycosides()` and `betalactams()`. 

In this post, we will explore how to use the `tidymodels` framework to predict resistance patterns in the `example_isolates` dataset in two examples. 

## Example 1: Using Antimicrobial Selectors

By leveraging the power of `tidymodels` and the `AMR` package, we’ll build a reproducible machine learning workflow to predict the Gramstain of the microorganism to two important antibiotic classes: aminoglycosides and beta-lactams.

### **Objective**

Our goal is to build a predictive model using the `tidymodels` framework to determine the Gramstain of the microorganism based on microbial data. We will:

1. Preprocess data using the selector functions `aminoglycosides()` and `betalactams()`.
2. Define a logistic regression model for prediction.
3. Use a structured `tidymodels` workflow to preprocess, train, and evaluate the model.

### **Data Preparation**

We begin by loading the required libraries and preparing the `example_isolates` dataset from the `AMR` package.

```{r}
# Load required libraries
library(AMR)          # For AMR data analysis
library(tidymodels)   # For machine learning workflows, and data manipulation (dplyr, tidyr, ...)

# Your data could look like this:
example_isolates

# Select relevant columns for prediction
data <- example_isolates %>%
  # select AB results dynamically
  select(mo, aminoglycosides(), betalactams()) %>%
  # replace NAs with NI (not-interpretable)
   mutate(across(where(is.sir),
                 ~replace_na(.x, "NI")),
          # make factors of SIR columns
          across(where(is.sir),
                 as.integer),
          # get Gramstain of microorganisms
          mo = as.factor(mo_gramstain(mo))) %>%
  # drop NAs - the ones without a Gramstain (fungi, etc.)
  drop_na()
```

**Explanation:**

- `aminoglycosides()` and `betalactams()` dynamically select columns for antimicrobials in these classes.
- `drop_na()` ensures the model receives complete cases for training.

### **Defining the Workflow**

We now define the `tidymodels` workflow, which consists of three steps: preprocessing, model specification, and fitting.

#### 1. Preprocessing with a Recipe

We create a recipe to preprocess the data for modelling.

```{r}
# Define the recipe for data preprocessing
resistance_recipe <- recipe(mo ~ ., data = data) %>%
  step_corr(c(aminoglycosides(), betalactams()), threshold = 0.9)
resistance_recipe
```

For a recipe that includes at least one preprocessing operation, like we have with `step_corr()`, the necessary parameters can be estimated from a training set using `prep()`:

```{r}
prep(resistance_recipe)
```

**Explanation:**

- `recipe(mo ~ ., data = data)` will take the `mo` column as outcome and all other columns as predictors.
- `step_corr()` removes predictors (i.e., antibiotic columns) that have a higher correlation than 90%.

Notice how the recipe contains just the antimicrobial selector functions - no need to define the columns specifically. In the preparation (retrieved with `prep()`) we can see that the columns or variables `r paste0("'", suppressMessages(prep(resistance_recipe))$steps[[1]]$removals, "'", collapse = " and ")` were removed as they correlate too much with existing, other variables.

#### 2. Specifying the Model

We define a logistic regression model since resistance prediction is a binary classification task.

```{r}
# Specify a logistic regression model
logistic_model <- logistic_reg() %>%
  set_engine("glm") # Use the Generalised Linear Model engine
logistic_model
```

**Explanation:**

- `logistic_reg()` sets up a logistic regression model.
- `set_engine("glm")` specifies the use of R's built-in GLM engine.

#### 3. Building the Workflow

We bundle the recipe and model together into a `workflow`, which organises the entire modeling process.

```{r}
# Combine the recipe and model into a workflow
resistance_workflow <- workflow() %>%
  add_recipe(resistance_recipe) %>% # Add the preprocessing recipe
  add_model(logistic_model) # Add the logistic regression model
resistance_workflow
```

### **Training and Evaluating the Model**

To train the model, we split the data into training and testing sets. Then, we fit the workflow on the training set and evaluate its performance.

```{r}
# Split data into training and testing sets
set.seed(123) # For reproducibility
data_split <- initial_split(data, prop = 0.8) # 80% training, 20% testing
training_data <- training(data_split) # Training set
testing_data <- testing(data_split)   # Testing set

# Fit the workflow to the training data
fitted_workflow <- resistance_workflow %>%
  fit(training_data) # Train the model
```

**Explanation:**

- `initial_split()` splits the data into training and testing sets.
- `fit()` trains the workflow on the training set.

Notice how in `fit()`, the antimicrobial selector functions are internally called again. For training, these functions are called since they are stored in the recipe.

Next, we evaluate the model on the testing data.

```{r}
# Make predictions on the testing set
predictions <- fitted_workflow %>%
  predict(testing_data)                # Generate predictions
probabilities <- fitted_workflow %>%
  predict(testing_data, type = "prob") # Generate probabilities

predictions <- predictions %>%
  bind_cols(probabilities) %>%
  bind_cols(testing_data) # Combine with true labels

predictions

# Evaluate model performance
metrics <- predictions %>%
  metrics(truth = mo, estimate = .pred_class) # Calculate performance metrics

metrics


# To assess some other model properties, you can make our own `metrics()` function
our_metrics <- metric_set(accuracy, kap, ppv, npv) # add Positive Predictive Value and Negative Predictive Value
metrics2 <- predictions %>%
  our_metrics(truth = mo, estimate = .pred_class) # run again on our `our_metrics()` function

metrics2
```

**Explanation:**

- `predict()` generates predictions on the testing set.
- `metrics()` computes evaluation metrics like accuracy and kappa.

It appears we can predict the Gram stain with a `r round(metrics$.estimate[1], 3) * 100`% accuracy based on AMR results of only aminoglycosides and beta-lactam antibiotics. The ROC curve looks like this:

```{r}
predictions %>%
  roc_curve(mo, `.pred_Gram-negative`) %>%
  autoplot()
```

### **Conclusion**

In this post, we demonstrated how to build a machine learning pipeline with the `tidymodels` framework and the `AMR` package. By combining selector functions like `aminoglycosides()` and `betalactams()` with `tidymodels`, we efficiently prepared data, trained a model, and evaluated its performance.

This workflow is extensible to other antimicrobial classes and resistance patterns, empowering users to analyse AMR data systematically and reproducibly.


---


## Example 2: Predicting AMR Over Time

In this second example, we aim to predict antimicrobial resistance (AMR) trends over time using `tidymodels`. We will model resistance to three antibiotics (amoxicillin `AMX`, amoxicillin-clavulanic acid `AMC`, and ciprofloxacin `CIP`), based on historical data grouped by year and hospital ward.

### **Objective**

Our goal is to:

1. Prepare the dataset by aggregating resistance data over time.
2. Define a regression model to predict AMR trends.
3. Use `tidymodels` to preprocess, train, and evaluate the model.

### **Data Preparation**

We start by transforming the `example_isolates` dataset into a structured time-series format.

```{r}
# Load required libraries
library(AMR)
library(tidymodels)

# Transform dataset
data_time <- example_isolates %>%
  top_n_microorganisms(n = 10) %>% # Filter on the top #10 species
  mutate(year = as.integer(format(date, "%Y")),  # Extract year from date
         gramstain = mo_gramstain(mo)) %>% # Get taxonomic names
  group_by(year, gramstain) %>%
  summarise(across(c(AMX, AMC, CIP), 
                   function(x) resistance(x, minimum = 0),
                   .names = "res_{.col}"), 
            .groups = "drop") %>% 
  filter(!is.na(res_AMX) & !is.na(res_AMC) & !is.na(res_CIP)) # Drop missing values

data_time
```

**Explanation:**
- `mo_name(mo)`: Converts microbial codes into proper species names.
- `resistance()`: Converts AMR results into numeric values (proportion of resistant isolates).
- `group_by(year, ward, species)`: Aggregates resistance rates by year and ward.

### **Defining the Workflow**

We now define the modeling workflow, which consists of a preprocessing step, a model specification, and the fitting process.

#### 1. Preprocessing with a Recipe

```{r}
# Define the recipe
resistance_recipe_time <- recipe(res_AMX ~ year + gramstain, data = data_time) %>%
  step_dummy(gramstain, one_hot = TRUE) %>%  # Convert categorical to numerical
  step_normalize(year) %>%  # Normalise year for better model performance
  step_nzv(all_predictors())  # Remove near-zero variance predictors

resistance_recipe_time
```

**Explanation:**
- `step_dummy()`: Encodes categorical variables (`ward`, `species`) as numerical indicators.
- `step_normalize()`: Normalises the `year` variable.
- `step_nzv()`: Removes near-zero variance predictors.

#### 2. Specifying the Model

We use a linear regression model to predict resistance trends.

```{r}
# Define the linear regression model
lm_model <- linear_reg() %>%
  set_engine("lm") # Use linear regression

lm_model
```

**Explanation:**
- `linear_reg()`: Defines a linear regression model.
- `set_engine("lm")`: Uses R’s built-in linear regression engine.

#### 3. Building the Workflow

We combine the preprocessing recipe and model into a workflow.

```{r}
# Create workflow
resistance_workflow_time <- workflow() %>%
  add_recipe(resistance_recipe_time) %>%
  add_model(lm_model)

resistance_workflow_time
```

### **Training and Evaluating the Model**

We split the data into training and testing sets, fit the model, and evaluate performance.

```{r}
# Split the data
set.seed(123)
data_split_time <- initial_split(data_time, prop = 0.8)
train_time <- training(data_split_time)
test_time <- testing(data_split_time)

# Train the model
fitted_workflow_time <- resistance_workflow_time %>%
  fit(train_time)

# Make predictions
predictions_time <- fitted_workflow_time %>%
  predict(test_time) %>%
  bind_cols(test_time) 

# Evaluate model
metrics_time <- predictions_time %>%
  metrics(truth = res_AMX, estimate = .pred)

metrics_time
```

**Explanation:**
- `initial_split()`: Splits data into training and testing sets.
- `fit()`: Trains the workflow.
- `predict()`: Generates resistance predictions.
- `metrics()`: Evaluates model performance.

### **Visualising Predictions**

We plot resistance trends over time for amoxicillin.

```{r}
library(ggplot2)

# Plot actual vs predicted resistance over time
ggplot(predictions_time, aes(x = year)) +
  geom_point(aes(y = res_AMX, color = "Actual")) +
  geom_line(aes(y = .pred, color = "Predicted")) +
  labs(title = "Predicted vs Actual AMX Resistance Over Time",
       x = "Year",
       y = "Resistance Proportion") +
  theme_minimal()
```

Additionally, we can visualise resistance trends in `ggplot2` and directly add linear models there:

```{r}
ggplot(data_time, aes(x = year, y = res_AMX, color = gramstain)) +
  geom_line() +
  labs(title = "AMX Resistance Trends",
       x = "Year",
       y = "Resistance Proportion") +
  # add a linear model directly in ggplot2:
  geom_smooth(method = "lm",
              formula = y ~ x,
              alpha = 0.25) +
  theme_minimal()
```

### **Conclusion**

In this example, we demonstrated how to analyze AMR trends over time using `tidymodels`. By aggregating resistance rates by year and hospital ward, we built a predictive model to track changes in resistance to amoxicillin (`AMX`), amoxicillin-clavulanic acid (`AMC`), and ciprofloxacin (`CIP`). 

This method can be extended to other antibiotics and resistance patterns, providing valuable insights into AMR dynamics in healthcare settings.
