---
title: "Data sets for download / own use"
date: '`r format(Sys.Date(), "%d %B %Y")`'
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 1
vignette: >
  %\VignetteIndexEntry{Data sets for download / own use}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE, results = "markup"}
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = TRUE,
  comment = "#",
  fig.width = 7.5,
  fig.height = 5
)

library(AMR)
library(dplyr)

options(knitr.kable.NA = "")

structure_txt <- function(dataset) {
  paste0(
    "A data set with ",
    format(nrow(dataset), big.mark = " "), " rows and ",
    ncol(dataset), " columns, containing the following column names:  \n",
    AMR:::vector_or(colnames(dataset), quotes = "*", last_sep = " and ", sort = FALSE), "."
  )
}

download_txt <- function(filename) {
  msg <- paste0(
    "It was last updated on ",
    trimws(format(file.mtime(paste0("../data/", filename, ".rda")), "%e %B %Y %H:%M:%S %Z", tz = "UTC")),
    ". Find more info about the contents, (scientific) source, and structure of this [data set here](https://msberends.github.io/AMR/reference/", ifelse(filename == "antivirals", "antimicrobials", filename), ".html).\n"
  )
  github_base <- "https://github.com/msberends/AMR/raw/main/"
  local_filename <- paste0("../data-raw/datasets/", filename)
  rds <- paste0(local_filename, ".rds")
  txt <- paste0(local_filename, ".txt")
  excel <- paste0(local_filename, ".xlsx")
  feather <- paste0(local_filename, ".feather")
  parquet <- paste0(local_filename, ".parquet")
  spss <- paste0(local_filename, ".sav")
  stata <- paste0(local_filename, ".dta")
  create_txt <- function(filename, type, software, exists) {
    if (isTRUE(exists)) {
      paste0(
        "* Download as [", software, "](", github_base, filename, ") (",
        AMR:::formatted_filesize(filename), ")  \n"
      )
    } else {
      paste0("* *(unavailable as ", software, ")*\n")
    }
  }

  if (any(
    file.exists(rds),
    file.exists(txt),
    file.exists(excel),
    file.exists(feather),
    file.exists(parquet),
    file.exists(spss),
    file.exists(stata)
  )) {
    msg <- c(
      msg, "\n**Direct download links:**\n\n",
      create_txt(rds, "rds", "original R Data Structure (RDS) file", file.exists(rds)),
      create_txt(txt, "txt", "tab-separated text file", file.exists(txt)),
      create_txt(excel, "xlsx", "Microsoft Excel workbook", file.exists(excel)),
      create_txt(feather, "feather", "Apache Feather file", file.exists(feather)),
      create_txt(parquet, "parquet", "Apache Parquet file", file.exists(parquet)),
      create_txt(spss, "sav", "IBM SPSS Statistics data file", file.exists(spss)),
      create_txt(stata, "dta", "Stata DTA file", file.exists(stata))
    )
  }
  paste0(msg, collapse = "")
}

print_df <- function(x, rows = 6) {
  x %>%
    as.data.frame(stringsAsFactors = FALSE) %>%
    head(n = rows) %>%
    mutate_all(function(x) {
      if (is.list(x)) {
        sapply(x, function(y) {
          if (length(y) > 3) {
            paste0(paste(y[1:3], collapse = ", "), ", ...")
          } else if (length(y) == 0 || all(is.na(y))) {
            ""
          } else {
            paste(y, collapse = ", ")
          }
        })
      } else {
        x
      }
    }) %>%
    knitr::kable(align = "c")
}
```

All reference data (about microorganisms, antimicrobials, SIR interpretation, EUCAST rules, etc.) in this `AMR` package are reliable, up-to-date and freely available. We continually export our data sets to formats for use in R, MS Excel, Apache Feather, Apache Parquet, SPSS, and Stata. We also provide tab-separated text files that are machine-readable and suitable for input in any software program, such as laboratory information systems. 

> If you are working in Python, be sure to use our [AMR for Python](https://msberends.github.io/AMR/articles/AMR_for_Python.html) package. It allows all relevant AMR data sets to be natively available in Python.

## `microorganisms`: Full Microbial Taxonomy

`r structure_txt(microorganisms)`

This data set is in R available as `microorganisms`, after you load the `AMR` package.

`r download_txt("microorganisms")`

**NOTE: The exported files for SPSS and Stata contain only the first 50 SNOMED codes per record, as their file size would otherwise exceed 100 MB; the file size limit of GitHub.** Their file structures and compression techniques are very inefficient. Advice? Use R instead. It's free and much better in many ways.

The tab-separated text file and Microsoft Excel workbook both contain all SNOMED codes as comma separated values.

**Example content**

Included (sub)species per taxonomic kingdom:

```{r, echo = FALSE}
microorganisms %>%
  count(kingdom) %>%
  mutate(n = format(n, big.mark = " ")) %>%
  setNames(c("Kingdom", "Number of (sub)species")) %>%
  print_df()
```

First 6 rows when filtering on genus *Escherichia*:

```{r, echo = FALSE}
microorganisms %>%
  filter(genus == "Escherichia") %>%
  print_df()
```


## `antimicrobials`: Antibiotic and Antifungal Drugs

`r structure_txt(antimicrobials)`

This data set is in R available as `antimicrobials`, after you load the `AMR` package.

`r download_txt("antimicrobials")`

The tab-separated text, Microsoft Excel, SPSS, and Stata files all contain the ATC codes, common abbreviations, trade names and LOINC codes as comma separated values.

**Example content**

```{r, echo = FALSE}
antimicrobials %>%
  filter(ab %in% colnames(example_isolates)) %>%
  print_df()
```


## `clinical_breakpoints`: Interpretation from MIC values & disk diameters to SIR

`r structure_txt(clinical_breakpoints)`

This data set is in R available as `clinical_breakpoints`, after you load the `AMR` package.

`r download_txt("clinical_breakpoints")`

**Example content**

```{r, echo = FALSE}
clinical_breakpoints %>%
  mutate(mo_name = mo_name(mo, language = NULL), .after = mo) %>%
  mutate(ab_name = ab_name(ab, language = NULL), .after = ab) %>%
  print_df()
```


## `microorganisms.groups`: Species Groups and Microbiological Complexes

`r structure_txt(microorganisms.groups)`

This data set is in R available as `microorganisms.groups`, after you load the `AMR` package.

`r download_txt("microorganisms.groups")`

**Example content**

```{r, echo = FALSE}
microorganisms.groups %>%
  print_df()
```


## `intrinsic_resistant`: Intrinsic Bacterial Resistance

`r structure_txt(intrinsic_resistant)`

This data set is in R available as `intrinsic_resistant`, after you load the `AMR` package.

`r download_txt("intrinsic_resistant")`

**Example content**

Example rows when filtering on *Enterobacter cloacae*:

```{r, echo = FALSE}
intrinsic_resistant %>%
  transmute(
    microorganism = mo_name(mo),
    antibiotic = ab_name(ab)
  ) %>%
  filter(microorganism == "Enterobacter cloacae") %>%
  arrange(antibiotic) %>%
  print_df(rows = Inf)
```


## `dosage`: Dosage Guidelines from EUCAST

`r structure_txt(dosage)`

This data set is in R available as `dosage`, after you load the `AMR` package.

`r download_txt("dosage")`

**Example content**

```{r, echo = FALSE}
dosage %>%
  print_df()
```


## `example_isolates`: Example Data for Practice

`r structure_txt(example_isolates)`

This data set is in R available as `example_isolates`, after you load the `AMR` package.

`r download_txt("example_isolates")`

**Example content**

```{r, echo = FALSE}
example_isolates %>%
  print_df()
```

## `example_isolates_unclean`: Example Data for Practice

`r structure_txt(example_isolates_unclean)`

This data set is in R available as `example_isolates_unclean`, after you load the `AMR` package.

`r download_txt("example_isolates_unclean")`

**Example content**

```{r, echo = FALSE}
example_isolates_unclean %>%
  print_df()
```


## `microorganisms.codes`: Common Laboratory Codes

`r structure_txt(microorganisms.codes)`

This data set is in R available as `microorganisms.codes`, after you load the `AMR` package.

`r download_txt("microorganisms.codes")`

**Example content**

```{r, echo = FALSE}
microorganisms.codes %>%
  print_df()
```


## `antivirals`: Antiviral Drugs

`r structure_txt(antivirals)`

This data set is in R available as `antivirals`, after you load the `AMR` package.

`r download_txt("antivirals")`

The tab-separated text, Microsoft Excel, SPSS, and Stata files all contain the trade names and LOINC codes as comma separated values.

**Example content**

```{r, echo = FALSE}
antivirals %>%
  print_df()
```
