---
title: "How to work with WHONET data"
author: "Matthijs S. Berends"
date: '`r format(Sys.Date(), "%d %B %Y")`'
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{How to work with WHONET data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE, results = 'markup'}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#",
  fig.width = 7.5,
  fig.height = 4.5
)
```

# Import of data

This tutorial assumes you already imported the WHONET data with e.g. the [`readxl` package](https://readxl.tidyverse.org/). In RStudio, this can be done using the menu button 'Import Dataset' in the tab 'Environment'. Choose the option 'From Excel' and select your exported file. Make sure date fields are imported correctly.

An example syntax could look like this:

```{r, eval = FALSE}
library(readxl)
data <- read_excel(path = "path/to/your/file.xlsx")
```

This package comes with an [example data set `WHONET`](./reference/WHONET.html). We will use it for this analysis.

# Preparation

First, load the relevant packages if you did not yet did this. I use the tidyverse for all of my analyses. All of them. If you don't know it yet, I suggest you read about it on their website: https://www.tidyverse.org/.

```{r, message = FALSE}
library(dplyr)   # part of tidyverse
library(ggplot2) # part of tidyverse
library(AMR)     # this package
```

We will have to transform some variables to simplify and automate the analysis:

* Microorganisms should be transformed to our own microorganism IDs (called an `mo`) using [the ITIS reference data set](./reference/ITIS.html), which contains all ~20,000 microorganisms from the taxonomic kingdoms Bacteria, Fungi and Protozoa. We do the tranformation with `as.mo()`. This function also recognises almost all WHONET abbreviations of microorganisms.
* Antimicrobial results or interpretations have to be clean and valid. In other words, they should only contain values `"S"`, `"I"` or `"R"`. That is exactly where the `as.rsi()` function is for.

```{r}
# transform variables
data <- WHONET %>%
  # get microbial ID based on given organism
  mutate(mo = as.mo(Organism)) %>% 
  # transform everything from "AMP_ND10" to "CIP_EE" to the new `rsi` class
  mutate_at(vars(AMP_ND10:CIP_EE), as.rsi)
```

No errors or warnings, so all values are transformed succesfully. Let's check it though, with a couple of frequency tables:

```{r, results = 'asis'}
# our newly created `mo` variable
data %>% freq(mo, nmax = 10)

# our transformed antibiotic columns
# amoxicillin/clavulanic acid (J01CR02) as an example
data %>% freq(AMC_ND2)
```

# Analysis

*(more will be available soon)*

